Modeling the impact of IT- and AI-related emissions in a university setting.

### 2025-07-31 doc/paper.tex
Modeling, top results, initial merge of stuff from it-emissions.tex

\documentclass[11pt]{article}
\usepackage[margin=0.7in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\assumption}[1]{{#1}}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{hyperref}
\usepackage[backend=biber,style=authoryear,autocite=inline]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

\addbibresource{ref.bib}
\title{UZH IT \& AI Emissions Analysis\vspace{-1em}}
\author{Massimiliano Poletto}
\date{\vspace{-1em}Last modified: 18.12.2024}

@@

\maketitle
\section{Introduction}

We are trying to answer two questions:
\begin{itemize}
    \item Are there emissions savings in moving UZH's IT infrastructure to the cloud?
    \item What is the impact of increased AI use (and, in particular, of LLMs) on UZH's emissions?
\end{itemize}

To this end, we need some UZH-specific information:
\begin{enumerate}
    \item The size of UZH's student and staff population and a rough breakdown of
        what kind of work they do (to estimate their use of IT and AI).
    \item Information about UZH's current IT infrastructure (number of servers,
        classroom equipment, etc.) and the energy use (electricity bills) of any
        campus data centers.
    \item An estimate of the cloud compute capacity required to host UZH's current IT requirements.
\end{enumerate}

In addition, we need information about the possible location of cloud data centers and
LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
of electricity in different regions.

We describe our modeling approach and our assumptions and uncertainties regarding these factors.

\section{Modeling approach}

At a high level, we model IT emissions primarily bottom-up, based on device counts and
estimates of their characteristics (embodied emissions, power utilization, and duty cycle).

The model can operate in two modes: deterministic, where each device type has a fixed set of
characteristics based on the mean of values we found in the literature, and stochastic, where
we pick device characteristics from truncated normal distributions with the mean and standard
deviation of the literature values. Of course, normality is a strong and probably incorrect
assumption, but it is a simple initial way to model the uncertainty in our estimates.

@@

\section{Data sources and unknowns}

\subsection{UZH population}

We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
counts and emissions factors for various types of IT equipment at UZH. It reports approximately
10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
hundred printers, copiers, and projectors.

By contrast, \textcite{uzhprofile} reports approximately 28,000 students and 7,300 staff
at UZH. Evidently the spreadsheet only lists university assets, not personal devices.

\assumption{In modeling overall university IT emissions as well as potential LLM use,
we assume that 90\% of the university population, 32,000 people, have a laptop and a mobile phone.}
We do not yet have any information about the breakdown of different functions / job roles, so
\assumption{we assume that IT use is uniform across the university}. Both of these assumptions
should be revisited.

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
121 & \textcite{ecoinvent}  \\ \hline
104 & \textcite{teehan2013} (low) \\ \hline
338 & \textcite{teehan2013} (high) \\ \hline
244 & \textcite{rarecoil} \\ \hline
202 & \textcite{unctadder2024} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:laptops}
\end{center}

The Ecoinvent numbers appear to be outliers on the low end.
The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
information for over 90 popular devices. Mean supply chain emissions are
244 kg CO$_2$e with a standard deviation of 128 kg.
\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
238 & \textcite{ecoinvent} \\ \hline
303 & \textcite{teehan2013} \\ \hline
403 & \textcite{unctadder2024} \\ \hline
289 & \textcite{dellpcf} \\ \hline
277 & \textcite{boavizta} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:desktops}
\end{center}

Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
Dell's Product Carbon Footprints (\textcite{dellpcf}), which they publish for each of their products.
A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
\assumption{We model desktop supply chain emissions using these statistics.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
383 & \textcite{teehan2013} \\ \hline
1252 & \textcite{davy2021} (Dell) \\ \hline
1912 & \textcite{davy2021} (EC2) \\ \hline
899 & \textcite{boavizta} (medium server)\\ \hline
\end{tabular}
\label{tab:embodied_emissions:servers}
\end{center}

\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
1912 kg with $\sigma = 885$ kg.
\assumption{We model server supply chain emissions using the statistics corresponding to Dell servers.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
50 & \textcite{unctadder2024} \\ \hline
50 & \textcite{lovehagen2023} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:phones}
\end{center}

@@

\begin{description}
    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
      (A typical data center rack contains 42 1U servers and one switch or router.)
\end{description}

@@

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).

@@

Energy consumption of LLMs is being studied extensively (\textcite{budennyy2022eco2ai},
\textcite{castano2023exploring},
\textcite{devries2023growing},
\textcite{gowda2024watt},
\textcite{harding2024watts}, 
\textcite{heguerte2023estimate},
\textcite{luccioni2022estimating},
\textcite{luccioni2023counting},
\textcite{patterson2021carbon}, 
\textcite{rodriguez2024evaluating},
\textcite{tripp2024measuring}).

@@

The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
However, \textcite{devries2023growing} provides the following information about ChatGPT:
\begin{itemize}
    \item Operational power consumption of 500 MWh / day.
    \item Approximately 1300 MWh consumed in training.
    \item Approximately 4000 servers.
\end{itemize}

@@

The Swiss Federal Office for the Environment published an extensive report (\textcite{krebs2018umweltbilanz})
on the environmental impact of electricity generation in Switzerland in 2018.

However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
hour block over the course of the year. This gives us twelve data points that describe average
carbon intensity variation over the course of the day.

\section{Missing data}

The following data would improve the accuracy of our model (in roughly decreasing order of utility):
\begin{itemize}
    \item Campus data center energy use (e.g., at the granularity of monthly electricity bills).
    \item Current campus data center workloads (to better model transition to the cloud).
    \item Academic calendar (e.g., how many days per year is the university full vs operating
        at lower levels).
    \item Common or otherwise representative pieces of equipment (servers, routers, standard desktops, etc.).
    \item Rough breakdown of UZH's population by field / role.
\end{itemize}

\section{Other considerations}

\subsection{Other IT components}

For now, we do not attempt to model changes in network infrastructure or utilization due to
move to the cloud or increased LLM use. See \textcite{jacob2023} for an overview of
network sustainability.

\subsection{Potential AI benefits}

We typically think of AI use as creating additional energy demands and GHG emissions.
So far, this model does not include potential energy savings caused by AI, such as improvements
to building energy management, campus logistics, etc.

\textcite{tomlinson2024carbon} make the argument that tasks such as writing
and illustrating can be done with fewer GHG emissions by AI than by humans. They compare the
estimated emissions of an LLM performing the task to those of the average human, calculated
by multiplying per-capita annual emissions by the time ($\approx 1$h) required to write or draw.
Using this approach, AI is orders of magnitude cheaper than humans, whether one uses per-capita
emissions from the US (15 t CO$_2$e/y) or India (1.9 t). Of course, this methodology focuses on
a specific creative act: the LLM does not reduce the human's overall annual emissions.

\printbibliography
\end{document}

>>>>

\documentclass[11pt]{article}
\usepackage[margin=0.7in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\assumption}[1]{{#1}}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting.}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}

@@

\maketitle

\section{Abstract}
Information technology (IT) plays major role in society and has a non-negligible carbon footprint: for example, at the University of Zürich, it accounted for over $6\%$ of total emissions in 2024. At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates. Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030. Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap. As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and break down the impact of IT and AI on the university's carbon footprint. In this paper we present a simple interactive emissions model that approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs. Our analysis shows that IT infrastructure's embodied emissions are a dominant component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.

\section{Introduction}

approximately $6.3\%$ of total emissions in 2024 \cite{uzh:sustainability:report},

artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet \cite{bick:ai:adoption}.

We consider only direct impacts of AI: the increased electricity consumption and resulting emissions. We omit indirect impacts of AI adoption, such as potential emissions reductions due to improved building management: at the moment they are too nebulous to predict.

We attempt to model full lifecycle impact, including manufacture, operation, and disposal. We include not only university-owned devices (servers, classroom PCs, etc.), but also student and staff personal devices, since they represent the bulk of on-campus devices.

Likewise, we attempt to model full life cycle impact for cloud infrastructure and for LLMs.

\section{Modeling methodology}

We model campus IT emissions bottom-up, estimating the number of different devices and their characteristics, including embodied emissions, maximum power use, duty cycle, and operational lifetime. The model can operate in a deterministic linear mode or in a stochastic mode that perturbs device parameters.

Likewise we model cloud data centers bottom-up, using publicly available emissions data and assuming a one-for-one replacement of on-campus servers with cloud instances of a certain size.

For LLMs, we estimate operational energy use (Wh per query) using recent public data. Non-operational emissions, including embodied emissions of the server fleet and the cost of training the models, are harder to estimate and attribute. Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.

@@

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
121 & \cite{ecoinvent}  \\ \hline
104 & \cite{teehan2013} (low) \\ \hline
338 & \cite{teehan2013} (high) \\ \hline
244 & \cite{rarecoil} \\ \hline
202 & \cite{unctadder2024} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:laptops}
\end{center}

The Ecoinvent numbers appear to be outliers on the low end.
The most comprehensive data comes from \cite{rarecoil}, which lists manufacturer
information for over 90 popular devices. Mean supply chain emissions are
244 kg CO$_2$e with a standard deviation of 128 kg.
\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
238 & \cite{ecoinvent} \\ \hline
303 & \cite{teehan2013} \\ \hline
403 & \cite{unctadder2024} \\ \hline
289 & \cite{dellpcf} \\ \hline
277 & \cite{boavizta} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:desktops}
\end{center}

Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
Dell's Product Carbon Footprints (\cite{dellpcf}), which they publish for each of their products.
A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
\assumption{We model desktop supply chain emissions using these statistics.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
383 & \cite{teehan2013} \\ \hline
1252 & \cite{davy2021} (Dell) \\ \hline
1912 & \cite{davy2021} (EC2) \\ \hline
899 & \cite{boavizta} (medium server)\\ \hline
\end{tabular}
\label{tab:embodied_emissions:servers}
\end{center}

\cite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
1912 kg with $\sigma = 885$ kg.
\assumption{We model server supply chain emissions using the statistics corresponding to Dell servers.}

@@

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
68 & \cite{googlepixel8} (Pixel 8)\\ \hline
54 & \cite{appleiphone13} (iPhone 13)\\ \hline
50 & \cite{unctadder2024} \\ \hline
50 & \cite{lovehagen2023} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:phones}
\end{center}

@@

\begin{description}
    \item[Computer monitors] 344 kg CO$_2$e (\cite{teehan2013}, \cite{dellpcf})
    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
    \item[Printer/copier stations] 1167 kg CO$_2$e (\cite{ecoinvent})
    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
      (A typical data center rack contains 42 1U servers and one switch or router.)
\end{description}

@@

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
infrastructure. The best resource we have been able to find is the Datavizta API (\cite{boavizta}).

@@

Energy consumption of LLMs is being studied extensively (\cite{budennyy2022eco2ai},
\cite{castano2023exploring},
\cite{devries2023growing},
\cite{gowda2024watt},
\cite{harding2024watts},
\cite{heguerte2023estimate},
\cite{luccioni2022estimating},
\cite{luccioni2023counting},
\cite{patterson2021carbon},
\cite{rodriguez2024evaluating},
\cite{tripp2024measuring}).

@@

The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
However, \cite{devries2023growing} provides the following information about ChatGPT:
\begin{itemize}
    \item Operational power consumption of 500 MWh / day.
    \item Approximately 1300 MWh consumed in training.
    \item Approximately 4000 servers.
\end{itemize}

@@

The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz})
on the environmental impact of electricity generation in Switzerland in 2018.

However, for the purposes of the model, we use data from \cite{electricitymaps}. Specifically,
we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
hour block over the course of the year. This gives us twelve data points that describe average
carbon intensity variation over the course of the day.

\subsection{Limitations and validation}

The biggest limitation is the modest availability of some ground truth data about UZH IT infrastructure. Specifically, we have only estimates of campus data center energy use, and limited information abut UZH research workloads (whether AI or not). We assume that many research AI workloads will be relatively similar in intensity to existing scientific computation workloads. We also have limited information (e.g., peak power consumption) about UZH use of external high-performance compute resources such as Alps/CSCS.

Additionally, cloud vendors and AI companies are secretive about their infrastructure. We have relied on public datasets and research literature to estimate cloud and AI emissions.

Despite these limitations, our bottom-up model aligns closely with two UZH data points. First, the model correctly estimates approximately $400$ kW of compute for central campus data centers. Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt CO$_2$e emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.

\section{Results}

Our analysis shows that manufacturing (“embodied”) emissions dominate UZH's IT carbon footprint, while operational impacts, including AI adoption, are relatively modest.

\begin{itemize}

\item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint. For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t CO$_2$e annually. Usage-related emissions amount to about a tenth of that.

\item Widespread adoption of LLMs and other frontier models should have moderate impact on total emissions. If everyone at UZH made heavy use of AI (15 queries daily), this would add approximately 220 t CO2e annually, of which 160 t directly attributable to queries. While this would almost double an individual's real-time energy use, the impact on overall emissions is relatively small.

\item Embodied emissions are an important consideration for university-owned devices as well. If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t CO$_2$e, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.

\item Data center cooling and efficiency represents a potential savings opportunity. The largest operational emissions are from data center servers and cooling (320 t). UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders. Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t CO$_2$e annually, though this estimate requires further validation.

\end{itemize}

\section{Related work}

\section{Conclusion}

\section{References}

\printbibliography

\end{document}
